\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-03-13 jasa/paper-jasa.tex
removing redundancies

Figures~\ref{fig:1c} and \ref{fig:1d} show the approximate $\log \left( \hat{r}(\mathbf{x}) \right)$ as a function of $\mathbf{x}$ using a 
2-layer neural network and a random forest for the classifier $\hat{s}(\mathbf{x})$.
%ratio when using respectively a linear model, a 2-layer neural network and a random
%forest. 
The blue curves corresponds to the exact log-ratio. The green curve shows 
$\log \left( (1-\hat s(\mathbf{x}))/\hat s(\mathbf{x}) \right)$
without calibration. The red curve is the approximation $\log \hat{r}(\mathbf{x})$ calibrated
using histograms. Finally, the
cyan curve shows the approximated log-likelihood ratio when decomposing the mixture, as
described in Section~\ref{sec:mixture},  and calibrating the underlying classifiers with histograms.

@@

The first fact to highlight is that, in the finite sample setting, uncalibrated
direct approximations $\log \left( \hat{r}(\mathbf{x}) \right)$, as shown in green, may not always yield very accurate
estimates of the ratio, even when using universal approximators such as a neural
network or a random forest. This observation is at the origin of this work, and
led us to show that the ratio $r(\mathbf{x})$ can still be computed exactly by
forming the equivalent statistic
$\sfrac{p(s(\mathbf{x})|\gamma=0.05)}{p(s(\mathbf{x})|\gamma=0)}$, provided the
transformation $s(\mathbf{x})$ is monotonic with $r(\mathbf{x})$. Indeed, when
constructing $\sfrac{\hat p(\hat s(\mathbf{x})|\gamma=0.05)}{\hat p(\hat
s(\mathbf{x})|\gamma=0)}$ as we propose, estimates appear much closer to the
true ratio than their direct counterparts, as illustrated in red. 
%The
%approximated ratios show to be more accurate for all three classifiers, even
%when the direct approximation was not very good in the first place, e.g. as for the
%linear model. 
In regions where the densities are low (i.e., at the
boundaries) the approximated ratio is less accurate since the underlying
classifier cannot make precise predictions due to the lack training samples
 in these parts of the input space. When leveraging the fact
that densities are mixtures and decomposing their ratio, the capacity of the
underlying classifiers can be more effectively focused on easier supervised learning tasks,
resulting as expected in even more accurate approximations, as shown in cyan.

As shown in the figures, qualitatively different results are obtained depending
the classification algorithm embedded within the approximation of the ratio. The
neural network-based approximation appears to work best, due to the capacity of
the classifier to model a smooth decision function $s(\mathbf{x})$ which is
very close to the optimal decision function. The random forest yields a decision function
which is piece-wise constant, therefore non-continuous and not strictly
monotonic with the ratio in the finite sample regime. On this problem, the
random forest still approximates the ratio correctly, but more variance can be
observed. 
%Finally, the linear model produces a decision function that is
%smooth and continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
%the ratio around $\mathbf{x}=1$, this still results in an accurate
%approximation, but poorer results should be expected on more complex problems.
While not directly apparent, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat
s(\mathbf{x}))$ is key to obtain accurate results. Standard
histograms with fixed bins have been used here for illustrative purposes, but we
anticipate that the embedded density estimation algorithm should be carefully
chosen and tuned, on a case by case basis. Overall, this first example confirms
that even if the classifier does a poor job at modeling the optimal decision
function, good approximations of the likelihood ratio can still be obtained
with the proposed method, by calibrating $s(\mathbf{x})$ and/or by decomposing
the mixture when possible.
%
%
%\begin{figure}
%    \centering
%    \begin{subfigure}[b]{0.3\textwidth}
%        \includegraphics[width=\textwidth]{figures/p_sij.png}
%        \caption{Pairwise classifier score distributions}
%        \label{fig:p_sij}
%    \end{subfigure}
%    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
%      %(or a blank line to force the subfigure onto a new line)
%    \begin{subfigure}[b]{0.3\textwidth}
%        \includegraphics[width=\textwidth]{figures/fig_ratio_dist.png}
%        \caption{Likelihood ratio distributions}
%        \label{fig:decomposed_ratio}
%    \end{subfigure}
%    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
%      %(or a blank line to force the subfigure onto a new line)
%    \begin{subfigure}[b]{0.3\textwidth}
%        \includegraphics[width=\textwidth]{figures/likelihood.png}
%        \caption{log-likelihood curve }
%        \label{fig:likelihood}
%    \end{subfigure}
%
%    \caption{Approximation of the log-likelihood ratio
%             $\log r(\mathbf{x};\gamma=0.05,\gamma=0) = \log \frac{p(\mathbf{x}|\gamma=0.05)}{p(\mathbf{x}|\gamma=0)}$
%             with calibrated classifiers.}
%\end{figure}
%
%%%%%%%%%%%%%

>>>>

Figures~\ref{fig:1c} and \ref{fig:1d} show the approximate $\log \left( \hat{r}(\mathbf{x}) \right)$ as a function of $\mathbf{x}$ using a 
2-layer neural network and a random forest for the classifier $\hat{s}(\mathbf{x})$.
The neural network provides a smoother approximation, while the random has some artifacts due to the fact that the
decision function is piece-wise constant.
%ratio when using respectively a linear model, a 2-layer neural network and a random
%forest. 
The blue curves show the exact log-ratio, the green curves show 
$\log \left( (1-\hat s(\mathbf{x}))/\hat s(\mathbf{x}) \right)$
without calibration, while the red curve is the improved approximation $\log \hat{r}(\mathbf{x})$ calibrated
using histograms. Finally, the cyan curve shows the approximated log-likelihood ratio when decomposing the mixture, as
seen in Fig.~\ref{fig:p_sij}.  By leveraging the fact that densities are mixtures, the capacity 
of the underlying classifiers can be more effectively focused on easier classification tasks,
resulting as expected in even more accurate approximations.

@@

%The first feature to highlight is that, in the finite sample setting, uncalibrated
%direct approximations $\log \left(  (1-\hat s(\mathbf{x}))/\hat s(\mathbf{x})  \right)$, as shown in green, may not yield a very accurate
%estimate of the ratio, even when using universal approximators such as a neural
%network or a random forest. 
%%This observation is at the origin of this work, and
%%led us to show that the ratio $r(\mathbf{x})$ can still be computed exactly by
%%forming the equivalent statistic
%%$\sfrac{p(s(\mathbf{x})|\gamma=0.05)}{p(s(\mathbf{x})|\gamma=0)}$, provided the
%%transformation $s(\mathbf{x})$ is monotonic with $r(\mathbf{x})$. 
%The approximation of the log-likelihood ratio is significantly improved when calibrating
%the classifier score with univariate density estimation
%% according to 
%%$\hat{r}(\mathbf{x}) = \sfrac{\hat p(\hat s(\mathbf{x})|\gamma=0.05)}{\hat p(\hat s(\mathbf{x})|\gamma=0)}$, 
%as illustrated in red. 
%%The
%%approximated ratios show to be more accurate for all three classifiers, even
%%when the direct approximation was not very good in the first place, e.g. as for the
%%linear model. 
%In regions where the densities are low (i.e., near the
%boundaries) the approximated ratio is less accurate.
%% since the underlying
%%classifier cannot make precise predictions due to the lack training samples
%% in these parts of the input space. 
% When leveraging the fact
%that densities are mixtures and decomposing their ratio as described in Section~\ref{sec:mixture}, 
%the capacity of the underlying classifiers can be more effectively focused on easier supervised learning tasks,
%resulting as expected in even more accurate approximations, as shown in cyan.

%As shown in the figures, qualitatively different results are obtained depending
%the classification algorithm embedded within the approximation of the ratio. The
%neural network-based approximation appears to work best, due to the capacity of
%the classifier to model a smooth decision function $s(\mathbf{x})$ which is
%very close to the optimal decision function. The random forest yields a decision function
%which is piece-wise constant, therefore non-continuous and not strictly
%monotonic with the ratio in the finite sample regime. On this problem, the
%random forest still approximates the ratio correctly, but more variance can be
%observed. 
%%Finally, the linear model produces a decision function that is
%%smooth and continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
%%the ratio around $\mathbf{x}=1$, this still results in an accurate
%%approximation, but poorer results should be expected on more complex problems.

While not directly apparent, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat
s(\mathbf{x}))$ is key to obtain accurate results. Standard
histograms with fixed bins have been used here for illustrative purposes, but we
anticipate that the embedded density estimation algorithm should be carefully
chosen and tuned, on a case by case basis. Overall, this first example confirms
that even if the classifier does a poor job at modeling the optimal decision
function, good approximations of the likelihood ratio can still be obtained
with the proposed method, by calibrating $s(\mathbf{x})$ and/or by decomposing
the mixture when possible.
%
%
%\begin{figure}
%    \centering
%    \begin{subfigure}[b]{0.3\textwidth}
%        \includegraphics[width=\textwidth]{figures/p_sij.png}
%        \caption{Pairwise classifier score distributions}
%        \label{fig:p_sij}
%    \end{subfigure}
%    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
%      %(or a blank line to force the subfigure onto a new line)
%    \begin{subfigure}[b]{0.3\textwidth}
%        \includegraphics[width=\textwidth]{figures/fig_ratio_dist.png}
%        \caption{Likelihood ratio distributions}
%        \label{fig:decomposed_ratio}
%    \end{subfigure}
%    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
%      %(or a blank line to force the subfigure onto a new line)
%    \begin{subfigure}[b]{0.3\textwidth}
%        \includegraphics[width=\textwidth]{figures/likelihood.png}
%        \caption{log-likelihood curve }
%        \label{fig:likelihood}
%    \end{subfigure}
%
%    \caption{Approximation of the log-likelihood ratio
%             $\log r(\mathbf{x};\gamma=0.05,\gamma=0) = \log \frac{p(\mathbf{x}|\gamma=0.05)}{p(\mathbf{x}|\gamma=0)}$
%             with calibrated classifiers.}
%\end{figure}
%
%%%%%%%%%%%%%
